---
title: Validation of Perceived Stress Scale
---
<style type="text/css">
   .main-container {max-width: 100%;}
   .row {display: flex;}
   .column {flex: 50%;}
</style>

<a id="top"></a>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, comment = NA, warning=FALSE, message=FALSE)
```

# **Project Overeview**

[<span style="font-size: 25px;"> Here is the Source Code1 </span>](https://github.com/WondimuAhmed/stress_scale)

The Perceived Stress Scale (PSS, 1) is a widely used tool for for assessing subjective stress. Despite its widespread application, empirical evidence regarding its psychometric properties within adolescent populations—particularly among Black youth—remains limited.  This study sought to fill that gap by evaluating the scale’s validity and reliability in this population. Specifically, it pursued three objectives: (1) to evaluate the applicability of the widely supported two-factor model(2,3) within a nationally representative sample of Black adolescents, (2) to examine its measurement invariance across gender, and (3) to assess its predictive validity by analyzing the relationship between perceived stress scores and lifetime prevalence of Major Depressive Disorder (MDD). 


```{r,load_packages,  echo=FALSE, warning=FALSE, message=FALSE}
if (!require("pacman")) install.packages("pacman");
pacman::p_load(tidyverse, eRm, lavaan, semTools, semPlot, psych, naniar, haven, reshape2, lavaanPlot, scales, flextable, DT,plotly, scales)

```

# **Methods**

## **Participants**

This study used data from the [National Survey of American Life – Adolescent Supplement (NSAL-A)](https://www.icpsr.umich.edu/web/RCMD/studies/36380). The sample included 1,170 Black adolescents aged 13–17 (52% female) representing the broader population of Black teens in the U.S. 

## **Measures**

### **Perceived Stress Scale (PSS)**

The study used the 14-item PSS from the NSAL-A, which included 8 positive and 6 negative items, with one item rephrased positively. Adolescents rated their perceived stress over the past month on a 5-point scale from 1 (never) to 5 (very often). To align with prior research, responses were recoded to a 0–4 scale.


### **Major Depressive Disorder (MDD)**

MDD was assessed using the World Mental Health Composite International Diagnostic Interview (CIDI) (4). 


## **Statistical Analyses**

### **Factor Structure** 

Confirmatory Factor Analysis (CFA) was performed to assess the factor structure of the PSS. We evaluated both one-factor and correlated two-factor models for the 14-item and 10-item versions. Model fit was evaluated using established indices: CFI and TLI (values > 0.95 indicate good fit), RMSEA (< 0.05 indicates good fit), and SRMR (< 0.05 indicates good fit). 

### **Measurement Invariance**

The study tested whether the structure of the PSS worked the same way for both males and females. We checked: 

- _Configural invariance_ (whether the scale measured stress similarly).
- _Metric invariance_ (whether the questions were understood the same way).
- _Scalar invariance_ (whether responses could be compared directly). 


### **Predictive Validity**

The study evaluated how well the PSS could identify adolescents with MDD using ROC curve analysis.

# **Data Preparation**

- Missing values coded as 8 and 9 in the database were replaced with NA to ensure accurate handling.
- Positively worded items were reverse-scored for consistency with negatively worded items.
- Descriptive statistics were calculated for key variables to detect anomaly.
- Correlation matrix was visualized through a heatmap to examine correlation patterns. 
- Missing data percentages were computed and visualized to identify variables with the most missing values. 

```{r echo=FALSE, warning=FALSE, message=FALSE}
adol_stress_data<- read.csv("Adolsstressnew.csv")
```

```{r replace_codes_en_reverse_score_items,  echo=FALSE}
# Replace codes 8 and 9 with NA 
adol_stress_data <- adol_stress_data %>%
  replace_with_na(replace = list(
    PC4D = c(8, 9),
    PC4E = c(8, 9),
    PC4G = c(8, 9),
    PC4H = c(8, 9),
    PC4I = c(8, 9),
    PC4J = c(8, 9),
    PC4N = c(8, 9)
  ))

# Reverse scoring positively worded items

adol_stress_data <- adol_stress_data %>%
  mutate(
    PC4B = 6 - PC4B,
    PC4D = 6 - PC4D,
    PC4E = 6 - PC4E,
    PC4F = 6 - PC4F,
    PC4G = 6 - PC4G,
    PC4I = 6 - PC4I,
    PC4J = 6 - PC4J,
    PC4M = 6 - PC4M
  )


```


```{r descriptive-stats-table, echo=FALSE}
# Create a descriptive statistics table for the Perceived Stress Scale items and related variables.
# This table includes the mean, standard deviation, median, minimum, and maximum values for each item.

desc_table<- adol_stress_data[, c(4,20:33)]
desc_table<- as.data.frame(describe(desc_table))%>% mutate(across(where(is.numeric), ~ round(., 2)))
desc_table <- desc_table[,c(3:5,8,9)]

# Rename columns for clarity in the output table
 desc_table <- desc_table %>%
  rename(
    `Mean` = mean,
    SD = sd,
    Median = median,
    `Min` = min,
    Max = max)

 # Display the formatted descriptive statistics table
datatable(
 desc_table,
  options = list(dom = 't'),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Table 1. Descriptive Statistics'
  )
)


```




```{r correlation-matrix-and-heat_map, echo=FALSE}

# Compute the correlation matrix
cor_matrix <- cor(adol_stress_data[, 20:33], use = "pairwise.complete.obs")


cor_melted <- melt(cor_matrix)
cor_melted <-as.data.frame(cor_melted )

# Plot the heatmap
corr_plot <- ggplot(data = cor_melted, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile(color = "white") +
  scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
                       midpoint = 0, limit = c(-1, 1), space = "Lab", 
                       name = "Correlation") +
                       theme_minimal() +
                       theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust = 1)) +
                       labs(title = "Figure 1. Heatmap of Correlations Among Perceived Stress Items",
                       x = "Items", y = "Items")

corr_plot 

```




```{r calculate-percentage-missing-and-plot, echo=FALSE}

# Specify the 14 items
items <- c("PC4A", "PC4B", "PC4C", "PC4D", "PC4E", "PC4F", "PC4G", 
           "PC4H", "PC4I", "PC4J", "PC4K", "PC4L", "PC4M", "PC4N")

# Calculate the percentage of missing values for each item
missing_percentages <- adol_stress_data %>%
  summarise(across(all_of(items), ~ mean(is.na(.)) * 100)) %>%
  pivot_longer(cols = everything(), names_to = "Item", values_to = "Missing_Percentage")

missing_percentages <- missing_percentages %>%
  arrange(desc(Missing_Percentage))

# Plot the missing percentages
ggplot(missing_percentages, aes(x = reorder(Item, Missing_Percentage), y = Missing_Percentage, fill = Missing_Percentage)) +
  geom_bar(stat = "identity", color = "black") +
  scale_fill_gradient(low = "blue", high = "red") +
  labs(
    title = "Figure 2. Percentage of Missing Values for Perceived Stress Items",
    x = "Items",
    y = "Missing Percentage (%)"
  ) + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +coord_flip()

```



# **Results**

## **Factor Structure**

The Tables below(Tables 2 and 3) summarizes the results of the goodness-of-fit tests for the one-factor and two-factor models of the PSS-14 and PSS-10. 


```{r unidimensional-model-pss14,  echo=FALSE, warning=FALSE, message=FALSE}
# Define the CFA model
pss_cfa1 <- '
  psg =~ PC4A + PC4C + PC4H + PC4K + PC4L + PC4N + 
         PC4B + PC4D + PC4E + PC4F + PC4G + 
         PC4I + PC4J + PC4M '

pss_fit1 <- cfa(pss_cfa1, data = adol_stress_data, std.lv = TRUE, 
              ordered = TRUE, mimic = c("MPlus"))

# summary(pss_fit1, standardized = TRUE, fit.measures = TRUE, rsquare = TRUE) 


#plots for model 1

labels = list(psg = "Gneneral Percieved Stress") 

p11 <- lavaanPlot(model = pss_fit1, labels = labels, graph_options = list(rankdir = "LR"), node_options = list(shape = "box", fontname = "Helvetica"), edge_options = list(color = "blue"), coefs = TRUE, covs = TRUE, stars = TRUE, digits = 2)


 # Extract parameter estimates
param_estimates <- parameterEstimates(pss_fit1, standardized = TRUE)

# Filter for standardized loadings and plot
std_loadings <- param_estimates %>%
  filter(op == "=~") %>% 
  dplyr::select(latent = lhs, indicator = rhs, std.loading = std.all)  

p12 <- ggplot(std_loadings, aes(x = reorder(indicator, std.loading), y = std.loading, fill = latent)) +
  geom_bar(stat = "identity", color = "black", width = .7) +
  coord_flip() + 
  labs(
    title = "Standardized Loadings for Unidimesional CFA Model",
    x = "Indicator",
    y = "Standardized Loading"
  ) +
   scale_colour_viridis_d ()+  
  theme_minimal() +scale_y_continuous(limits = c(-0.5, 1)) +
  theme(
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5), legend.position = "none",
  )

```



```{r, results='hide', fig.show='hide', echo=FALSE}
p12
```


```{r results='hide', fig.show='hide',echo=FALSE}
p11
```

```{r , two-factor-correlated-model-pss14, echo=FALSE, warning=FALSE, message=FALSE}

# Define the second CFA model
pss_cfa2 <- '
  ps1 =~ PC4A + PC4C + PC4H + PC4K + PC4L + PC4N
  ps2 =~ PC4B + PC4D + PC4E + PC4F + PC4G + PC4I + PC4J + PC4M '

pss_fit2 <- cfa(pss_cfa2, data = adol_stress_data, std.lv = TRUE, 
              ordered = TRUE, mimic = c("MPlus"))

labels = list(ps1 = "Percieved Coping",ps2 = "Percieved Distress" ) 

p21 <- lavaanPlot(model = pss_fit2, labels = labels, graph_options = list(rankdir = "LR"), node_options = list(shape = "box", fontname = "Helvetica"), edge_options = list(color = "blue"), coefs = TRUE, covs = TRUE, stars = TRUE, digits = 2)

# Extract parameter estimates
param_estimates2 <- parameterEstimates(pss_fit2, standardized = TRUE)

# Filter for standardized loadings and plot
std_loadings <- param_estimates2 %>%
  filter(op == "=~") %>%   
  dplyr::select(latent = lhs, indicator = rhs, std.loading = std.all)  

p22 <- ggplot(std_loadings, aes(x = reorder(indicator, std.loading), y = std.loading, fill = latent)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  coord_flip() +  
  labs(
    title = "Standardized Loadings for Two Factor CFA Model",
    x = "Indicator",
    y = "Standardized Loading"
  ) +
  scale_fill_viridis_d(option = "plasma")   +  
  theme_minimal() + scale_y_continuous(limits = c(-.5, 1)) +  # Set y-axis limits to 0 and 
  theme(
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5), legend.position = "none",
  )
```


  
```{r results='hide',fig.show='hide', echo=FALSE}
p22

```

  
``` {r results='hide', fig.show='hide', echo=FALSE}
p21
```

```{r, two-factor-correlated-with-13-items, echo=FALSE, warning=FALSE, message=FALSE}

# Define the revised CFA model

pss_cfa2a <- '
  ps1 =~ PC4A + PC4C + PC4H + PC4K + PC4N
  ps2 =~ PC4B + PC4D + PC4E + PC4F + PC4G + PC4I + PC4J + PC4M '

pss_fit2a <- cfa(pss_cfa2a, data = adol_stress_data, std.lv = TRUE, 
               ordered = TRUE, mimic = c("MPlus"))

labels = list(ps1 = "Percieved Coping",ps2 = "Percieved Distress" )

p31 <- lavaanPlot(model = pss_fit2a, labels = labels, graph_options = list(rankdir = "LR"), node_options = list(shape = "box", fontname = "Helvetica"), edge_options = list(color = "blue"), coefs = TRUE, covs = TRUE, stars = TRUE, digits = 2)


# Extract parameter estimates
param_estimates2 <- parameterEstimates(pss_fit2a, standardized = TRUE)

# Filter for standardized loadings and plot

std_loadings <- param_estimates2 %>%
  filter(op == "=~") %>%  
  dplyr::select(latent = lhs, indicator = rhs, std.loading = std.all)  

p32 <- ggplot(std_loadings, aes(x = reorder(indicator, std.loading), y = std.loading, fill = latent)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  coord_flip() +  
  labs(
    title = "Standardized Loadings for Two Factor CFA Model",
    x = "Indicator",
    y = "Standardized Loading"
  ) +
  scale_fill_viridis_d(option = "plasma") +  
  theme_minimal() + scale_y_continuous(limits = c(0, 1)) +
  theme(
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5), legend.position = "none",
  )
```

  
```{r fig.show='hide', echo=FALSE}
p32

```


  
``` {r echo=FALSE}
# htmltools::tagList(
#   htmltools::tags$h4("Figure 3. Path Diagram of the Two-Factor Model (PSS-13)"),
#   p31
# )

```




```{r model-comparison-pss14,  echo=FALSE, warning=FALSE, message=FALSE}

# Compare multiple CFA models using compareFit; compareFit is used to evaluate the relative fit of these models.

myout.mod1_3 <- compareFit(One_factor_model_14itms=pss_fit1, Two_factor_model_14itms=pss_fit2, Two_factor_model_13itms=pss_fit2a, nested = FALSE)
   
model_fit_indices1 <- as.data.frame(myout.mod1_3@fit)
model_fit_indices1  <- model_fit_indices1 [, c("cfi.robust", "tli.robust", "rmsea.robust", "srmr")] %>% mutate(across(where(is.numeric), ~ round(., 3)))
model_fit_indices1 <- model_fit_indices1  %>%
  rename(
    "CFI" = "cfi.robust",
    "TLI" = "tli.robust",
    "RMSEA" = "rmsea.robust",
    "SRMR" = "srmr") 

rownames(model_fit_indices1 ) <- c("Two Factor Model-13 Items", "Two Factor Model-14 Items", "One Factor Model-14 Items")   

 datatable(
  model_fit_indices1,
  options = list(dom = 't'),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Table 2. Model Fit Indices for Competing CFA Models (PSS-14)'
  )
)



```


```{r,scoring-pss13, echo=FALSE}
# Calculate scores for overall pss and subscales 

adol_stress_data <- adol_stress_data %>%
  mutate(
    PC4BS = PC4B - 1,  
    PC4DS = PC4D - 1, 
    PC4ES = PC4E - 1,  
    PC4FS = PC4F - 1, 
    PC4GS = PC4G - 1,  
    PC4IS = PC4I - 1,  
    PC4JS = PC4J - 1,  
    PC4MS = PC4M - 1,  

    PC4AS = PC4A - 1,    
    PC4CS = PC4C - 1,  
    PC4HS = PC4H - 1,    
    PC4KS = PC4K - 1,    
    PC4NS = PC4N - 1    
  )

# Calculate subscale scores
adol_stress_data <- adol_stress_data %>%
  mutate(
    # Perceived stress (negative) subscale
    PSS13N = PC4AS + PC4CS + PC4HS + PC4KS + PC4NS,

    # Perceived stress (positive) subscale
    PSS13P = PC4BS + PC4DS + PC4ES + PC4FS + PC4GS + 
             PC4IS + PC4JS + PC4MS,

    # Total perceived stress score
    PSS13Total = PSS13N + PSS13P
  )

```


```{r unidimensional-model-pss10 , echo=FALSE, warning=FALSE, message=FALSE}
# Define unidimensional CFA for pss-10

pss_cfa1b <-'psg=~ PC4A  + PC4C + PC4H  + PC4K + PC4N+ PC4B + PC4F + PC4G  + PC4I  + PC4J'
   
pss_fit1b <- cfa(pss_cfa1b, data=adol_stress_data, std.lv=TRUE,ordered = TRUE, mimic =c("MPlus") )
#summary(pss_fit1b,standardized=TRUE,fit.measures=TRUE, rsquare=TRUE)

labels = list(psg = "Gneneral Percieved Stress") 

p41 <- lavaanPlot(model = pss_fit1b, labels = labels, graph_options = list(rankdir = "LR"), node_options = list(shape = "box", fontname = "Helvetica"), edge_options = list(color = "blue"), coefs = TRUE, covs = TRUE, stars = TRUE, digits = 2)


# Extract parameter estimates
param_estimates <- parameterEstimates(pss_fit1b, standardized = TRUE)

# Filter for standardized loadings and plot

std_loadings <- param_estimates %>%
  filter(op == "=~") %>%   # Select only factor loadings
  dplyr::select(latent = lhs, indicator = rhs, std.loading = std.all)  

p42 <- ggplot(std_loadings, aes(x = reorder(indicator, std.loading), y = std.loading, fill = latent)) +
  geom_bar(stat = "identity", color = "black", width = 0.7) +
  coord_flip() +  
  labs(
    title = "Standardized Loadings for Unidimesional CFA Model",
    x = "Indicator",
    y = "Standardized Loading"
  ) +
  scale_colour_viridis_d () +  
  theme_minimal() +scale_y_continuous(limits = c(0, 1)) +
  theme(
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5), legend.position = "none",
  )

```


  
```{r fig.show='hide', echo=FALSE}
p42
```


  
```{r fig.show='hide', results='hide', echo=FALSE}
p41
```

  

```{r correlated-two-factor-pss10, echo=FALSE, warning=FALSE, message=FALSE}
#Deffine the second model for pss10

pss_cfa3b <-'ps1=~PC4A  + PC4C + PC4H  + PC4K + PC4N   
          ps2=~ PC4B + PC4F + PC4G  + PC4I  + PC4J '

pss_fit3b <- cfa(pss_cfa3b, data=adol_stress_data, std.lv=TRUE,ordered=TRUE, mimic =c("MPlus"))
# summary(pss_fit3b,standardized=TRUE,fit.measures=TRUE, rsquare=TRUE)
#lavInspect(pss_fit3b, "cor.lv")

labels = list(ps1 = "Percieved Coping",ps2 = "Percieved Distress" ) 

p51 <- lavaanPlot(model = pss_fit3b, labels = labels, graph_options = list(rankdir = "LR"), node_options = list(shape = "box", fontname = "Helvetica"), edge_options = list(color = "blue"), coefs = TRUE, covs = TRUE, stars = TRUE, digits = 2)


# Extract parameter estimates
param_estimates2 <- parameterEstimates(pss_fit3b, standardized = TRUE)

# Filter for standardized loadings and plot

std_loadings <- param_estimates2 %>%
  filter(op == "=~") %>%   # Select only factor loadings
  dplyr::select(latent = lhs, indicator = rhs, std.loading = std.all)  

p52 <- ggplot(std_loadings, aes(x = reorder(indicator, std.loading), y = std.loading, fill = latent)) +
  geom_bar(stat = "identity", color = "blue", width = 0.7) +
  coord_flip() +  
  labs(
    title ="Standardized Loadings Two Factor PSS10 Model",
    x = "Indicator",
    y = "Standardized Loading"
  ) +
  scale_fill_viridis_d(option = "plasma") +  
  theme_minimal() + scale_y_continuous(limits = c(0, 1)) +  # Set y-axis limits to 0 and 
  theme(
    axis.text.x = element_text(size = 10, face = "bold"),
    axis.text.y = element_text(size = 10, face = "bold"),
    legend.title = element_text(size = 12, face = "bold"),
    legend.text = element_text(size = 10),
    plot.title = element_text(size = 14, face = "bold", hjust = 0.5), legend.position = "none",
  )
```


  
```{r fig.show='hide',echo=FALSE}
p52

```


 

  
```{r model-comparison-pss10, echo=FALSE, warning=FALSE, message=FALSE}

 models_1en2 <- compareFit(One_factor_model_10itms=pss_fit1b, Two_factor_model_10itms=pss_fit3b, nested = FALSE)
   #summary(models_1en2)
  
   
model_fit_indices <- as.data.frame(models_1en2@fit)
model_fit_indices  <- model_fit_indices [, c("cfi.robust", "tli.robust", "rmsea.robust", "srmr")] %>% mutate(across(where(is.numeric), ~ round(., 3)))

model_fit_indices <- model_fit_indices  %>%
  rename(
    "CFI" = "cfi.robust",
    "TLI" = "tli.robust",
    "RMSEA" = "rmsea.robust",
    "SRMR" = "srmr") 

rownames(model_fit_indices ) <- c("Two Factor Model-10 Items", "One Factor Model-10 Items")   

datatable(
  model_fit_indices,
  options = list(dom = 't'),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Table 3. Model Fit Indices for Competing CFA Models (PSS-10)'
  )
)

```

The one-factor models for both versions showed poor fit, while the two-factor models fit the data significantly better. In the PSS-14, item 12 had a very low factor loading (.093) on the perceived distress factor and was removed, resulting in a revised PSS-13 model. Both the revised PSS-13 and the PSS-10 two-factor models demonstrated good fit, supporting the theory that stress consists of two dimensions: perceived coping and perceived distress (see Figures 3 & 4)

  
``` {r path-diagram-2-factor-pss13, echo=FALSE}
htmltools::tagList(
  htmltools::tags$h4("Figure 3. Path Diagram of the Two-Factor Model (PSS-13)"),
  p31
)

```


 
``` {r path-diagram-2-factor-pss10, echo=FALSE}
htmltools::tagList(
  htmltools::tags$h4("Figure 4. Path Diagram of the Two-Factor Model (PSS-10)"),
  p51
)

```

```{r, scoring_pss10, echo=FALSE, warning=FALSE, message=FALSE}

# Scoring PSS10 
adol_stress_data <- adol_stress_data %>%
  mutate(
    PSS10N = PC4AS + PC4CS + PC4HS + PC4KS + PC4NS,
    PSS10P = PC4BS + PC4FS + PC4GS + PC4IS + PC4JS,
    PSS10Total = PSS10N + PSS10P
  )

```


## **Measurement Invariance**

The study found that the structure of the Perceived Stress Scale (PSS) is consistent and works equally well for both males and females, ensuring fairness in its use ( See Tables 4 & 5). However, girls reported significantly higher stress levels than boys on both versions of the scale. This highlights the importance of understanding and addressing gender differences when evaluating and managing stress in adolescents.

```{r gender_measurement_equivalence_pss13, echo=FALSE, warning=FALSE, message=FALSE}

# Ensure gender is treated as a factor
adol_stress_data$Gender <- as.factor(adol_stress_data$SEX)

# Configural Model
pss13_config <- cfa(
  pss_cfa2a, 
  data = adol_stress_data, 
  group = "Gender", 
  ordered = TRUE
)

#  Metric Invariance Model
pss13_metric <- cfa(
  pss_cfa2a, 
  data = adol_stress_data, 
  group = "Gender", 
  ordered = TRUE, 
  group.equal = "loadings"
)

# Scalar Invariance Model
pss13_scalar <- cfa(
  pss_cfa2a, 
  data = adol_stress_data, 
  group = "Gender", 
  ordered = TRUE, 
  group.equal = c("thresholds", "loadings")
)


# Model Comparison: Likelihood Ratio Test
lrt_results <- lavTestLRT(
  pss13_config, 
  pss13_metric, 
  pss13_scalar
)



# Model Comparison: Fit Indices
fit_comparison <- compareFit(
  pss13_config, 
  pss13_metric, 
  pss13_scalar
)


fit_comparison1 <- as.data.frame(fit_comparison@fit)
fit_comparison1   <- fit_comparison1  [, c("cfi", "tli", "rmsea", "srmr")] %>% mutate(across(where(is.numeric), ~ round(., 3)))

fit_comparison1  <- fit_comparison1  %>%
  rename(
    "CFI" = "cfi",
    "TLI" = "tli",
    "RMSEA" = "rmsea",
    "SRMR" = "srmr") 

rownames(fit_comparison1) <- c("Configural", "Metric", "Scalar") 

#datatable(fit_comparison1,options = list(dom = 't'))

datatable(
  fit_comparison1,
  options = list(dom = 't'),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Table 4. Model Fit Indices for Configural, Metric, and Scalar Invariance Models (PSS-14)'
  )
)


fit_diff1 <- as.data.frame(fit_comparison@fit.diff)
fit_diff1    <- fit_diff1[, c("cfi", "tli", "rmsea", "srmr")] %>% mutate(across(where(is.numeric), ~ round(., 3)))

fit_diff1 <- fit_diff1  %>%
  rename(
    "ΔCFI" = "cfi",
    "ΔTLI" = "tli",
    "ΔRMSEA" = "rmsea",
    "ΔSRMR" = "srmr") 

rownames(fit_diff1 ) <- c("Metric Vs.Configural", "Scalar Vs.Metric") 
datatable(fit_diff1,options = list(dom = 't')) 


```

```{r gender_measurement_equivalence_pss10,  echo=FALSE, warning=FALSE, message=FALSE}


# Configural
pss10_config <- cfa(
  model = pss_cfa3b, 
  data = adol_stress_data, 
  group = "Gender", 
  ordered = TRUE
)

# Metric
pss10_metric <- cfa(
  model = pss_cfa3b, 
  data = adol_stress_data, 
  group = "Gender", 
  ordered = TRUE, 
  group.equal = "loadings"
)

# Scalar 
pss10_scalar <- cfa(
  model = pss_cfa3b, 
  data = adol_stress_data, 
  group = "Gender", 
  ordered = TRUE, 
  group.equal = c("thresholds", "loadings")
)



# Likelihood Ratio Test for Nested Models
lrt_results <- lavTestLRT(
  pss10_config, 
  pss10_metric, 
  pss10_scalar
)
#print(lrt_results)

# Fit Index Comparison for Nested Models
fit_comparison <- compareFit(
  pss10_config, 
  pss10_metric, 
  pss10_scalar
)
#summary(fit_comparison)

fit_comparison2 <- as.data.frame(fit_comparison@fit)
fit_comparison2   <- fit_comparison2  [, c("cfi", "tli", "rmsea", "srmr")] %>% mutate(across(where(is.numeric), ~ round(., 3)))
fit_comparison2  <- fit_comparison2  %>%
  rename(
    "CFI" = "cfi",
    "TLI" = "tli",
    "RMSEA" = "rmsea",
    "SRMR" = "srmr") 

rownames(fit_comparison2) <- c("Configural", "Metric", "Scalar") 


datatable(
  fit_comparison2,
  options = list(dom = 't'),
  caption = htmltools::tags$caption(
    style = 'caption-side: top; text-align: left; font-weight: bold;',
    'Table 5. Model Fit Indices for Configural, Metric, and Scalar Invariance Models (PSS-10)'
  )
)


fit_diff2 <- as.data.frame(fit_comparison@fit.diff)
fit_diff2    <- fit_diff2[, c("cfi", "tli", "rmsea", "srmr")] %>% mutate(across(where(is.numeric), ~ round(., 3)))

fit_diff2 <- fit_diff2  %>%
  rename(
    "ΔCFI" = "cfi",
    "ΔTLI" = "tli",
    "ΔRMSEA" = "rmsea",
    "ΔSRMR" = "srmr") 

rownames(fit_diff2 ) <- c("Metric Vs.Configural", "Scalar Vs.Metric") 
datatable(fit_diff2,options = list(dom = 't')) 



```




##  **Predictive Validity**


The Perceived Stress Scale (PSS) showed moderate success in identifying teens at risk for depression(See Roc Curves below). Among the study group, 7.5% were diagnosed with depression, and the PSS-13 and PSS-10 versions both performed reasonably well in distinguishing these individuals. The "perceived distress" part of the scale was particularly effective, highlighting its importance in identifying teens who may need mental health support. This underscores the value of stress assessments in early detection and potential intervention for mental health challenges.

```{r, create_roc_curve, echo=FALSE}
# Create ROC curves
adol_stress_data$DSM_MDD <- as.factor(adol_stress_data$DSM_MDD)

adol_stress_data_d1 <- adol_stress_data[, c(3, 47:49)]
adol_stress_data_d1 <- na.omit(adol_stress_data_d1)

# Create basic ROC objects for Total, Negative, and Positive subscales
rocobj1 <- pROC::roc(adol_stress_data_d1$DSM_MDD, adol_stress_data_d1$PSS13Total, print.auc = TRUE)
rocobj2 <- pROC::roc(adol_stress_data_d1$DSM_MDD, adol_stress_data_d1$PSS13P)
rocobj3 <- pROC::roc(adol_stress_data_d1$DSM_MDD, adol_stress_data_d1$PSS13N)

# Combine ROC objects into a list
roclist <- list(
  "Total Score" = rocobj1,       
  "Perceived Coping" = rocobj2,  
  "Perceived Distress" = rocobj3 
)

g_roc1 <- pROC::ggroc(roclist, legacy.axes = TRUE) +
  geom_abline(linetype = "dashed", color = "gray") + # Add a dashed diagonal line
  geom_line(size = 1.5) +                           # Make ROC curves bold
  theme_classic() +
  ggtitle("Predictive Accuracy of PSS-13") +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate",
    linetype = "PSS-10"
  ) +
  theme(
    text = element_text(size = 14, face = "bold"),  # Larger, bold font for all text
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # Center and bold title
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles
    axis.text = element_text(size = 12),  # Larger font size for axis ticks
    legend.title = element_blank(),  # Bold legend title
    legend.text = element_text(size = 12)
  # Larger font for legend text
  ) +
  scale_color_manual(values = c("blue", "red", "green"))  # Custom colors for ROC curves

# Save the plot
roc_plot1 <- g_roc1


adol_stress_data_d2 <- adol_stress_data[, c(3, 47:53)]
adol_stress_data_d2 <- na.omit(adol_stress_data_d2)

# Create a basic roc object
rocobj1 <- pROC::roc(adol_stress_data_d2$DSM_MDD, adol_stress_data_d2$PSS10Total)
rocobj2 <- pROC::roc(adol_stress_data_d2$DSM_MDD, adol_stress_data_d2$PSS10P)
rocobj3 <- pROC::roc(adol_stress_data_d2$DSM_MDD, adol_stress_data_d2$PSS10N)
roclist1 <- list("Total Score" = rocobj1,
                "Perceived Coping" = rocobj2,
                "Perceived Distress" = rocobj3)

# Create the enhanced ROC plot
g_roc2 <- pROC::ggroc(roclist1, legacy.axes = TRUE) +
  geom_abline(linetype = "dashed", color = "gray") + # Add a dashed diagonal line
  geom_line(size = 1.5) +                           # Make ROC curves bold
  theme_classic() +
  ggtitle("Predictive Accuracy of PSS-10") +
  labs(
    x = "False Positive Rate",
    y = "True Positive Rate",
    linetype = "PSS-10"
  ) +
  theme(
    text = element_text(size = 14, face = "bold"),  # Larger, bold font for all text
    plot.title = element_text(size = 16, face = "bold", hjust = 0.5),  # Center and bold title
    axis.title = element_text(size = 14, face = "bold"),  # Bold axis titles
    axis.text = element_text(size = 12),  # Larger font size for axis ticks
    legend.title = element_blank(),  # Bold legend title
    legend.text = element_text(size = 12)
  # Larger font for legend text
  ) +
  scale_color_manual(values = c("blue", "red", "green"))  # Custom colors for ROC curves

# Save the plot
roc_plot2 <- g_roc2


```

<div class = "row">
<div class = "column">
    
```{r echo=FALSE}
 roc_plot1
```
</div>

<div class = "column">
```{r echo=FALSE}
  roc_plot2
```

</div>
</div>

# **Impications** 
 
These findings have important implications for both researchers and clinicians working with this population: 

- Use of the PSS-13 (excluding item 12) is recommended for Black adolescent populations, as it offers better model fit and construct representation compared to the full scale.
- Clinical screening efforts should prioritize the Perceived Distress subscale, which has greater relevance for identifying adolescents at risk for depression.
- Researchers and clinicians should avoid relying solely on the total PSS score, as it may obscure important distinctions between distress and coping and fails to capture the multidimensional nature of perceived stress in this population.
- The two-factor structure of the PSS underscores that perceived stress is not a unidimensional construct for Black adolescents; using a summed total score may lead to misinterpretation of their stress experiences.
- The PSS demonstrates strong reliability and validity among Black adolescents, supporting its continued use in research and clinical assessments within this group.
- Measurement invariance across gender supports equitable interpretation of the PSS-13 scores, making it an appropriate tool for use with both male and female Black adolescents.
- These findings call for culturally responsive and psychometrically sound approaches to stress assessment, encouraging the use of validated subscales to guide intervention and deepen understanding of stress-related outcomes in Black youth.

# **References**

1. Cohen, S., Kamarch, T., & Mermelstein, R. (1983). A global measure of perceived stress. Journal of Health and Social Behavior, 24, 385–396.

2. Hewitt, P. L., Flett, G. L., & Mosher, S. W. (1992). The perceived stress scale: Factor structure and relation to depression symptoms in a psychiatric sample. Journal of Psychopathology and Behavioral Assessment, 14(3), 247–257. https://doi.org/10.1007/BF00962631

3. Roberti, J. W., Harrington, L. N., & Storch, E. A. (2006). Further psychometric support for the 10-item version of the perceived stress scale. Journal of College Counseling, 9(2), 135–147. https://doi.org/10.1002/j.2161-1882.2006.tb00100.x

4. Wittchen, H. U. (1994). Reliability and validity studies of the WHO–Composite International Diagnostic Interview (CIDI): A critical review. Journal of Psychiatric Research, 28(1), 57–84. https://doi.org/10.1016/0022-3956(94)90036-1

5. Arnett, J. J. (1999). Adolescent storm and stress, reconsidered. The American Psychologist, 54(5), 317–326. https://doi.org/10.1037//0003-066x.54.5.317




<a href="#page-title" style="font-size: 20px; font-weight: bold; color: white; position: fixed; right: 20px; bottom: 20px; background-color: red; padding: 10px 15px; border-radius: 5px; box-shadow: 2px 2px 5px rgba(0, 0, 0, 0.2); text-decoration: none;">
    Back to Top
</a>



<a href="https://wondimuahmed.github.io/Portfolio/" style="font-size: 20px; font-weight: bold; color:white; position: fixed; bottom: 20px; background-color: #007BFF; padding: 10px 15px; border-radius: 5px; box-shadow: 2px 2px 5px rgba(0, 0, 0, 0.2); text-decoration: none;">
    Back to Portfolio
</a>


